Narrative

### 2023-05-21 manuscript/methods.qmd
method: fisher z + model sections improved

Perhaps the most common way to meta-analyze correlations is to first transform them into Fisher $z$ correlations (Borenstein et al., 2009, p. 41 - 43). The Fisher $z$ transformation is the inverse hyperbolic tangent of the product moment correlation and can be computed as $z = 0.5 \times ln\left(\frac{1+r}{1-r}\right)$ and its variance approximated as $V_z = 1 / (n -3)$. There is controversy in the use of the Fisher $z$ transformation for meta-analysis (e.g., Hunter & Schmidt, 2015, p. XX, ch. 5 "Use of Fisher’s z in Meta-Analysis of Correlations"), largely related to whether Pearson's $r$ or Fisher's $z$ leads to less bias in average correlations (Schulze, 2007). We observe that estimating heterogeneity does not seem to have been the focus of the literature's discussion of the appropriateness of the Fisher $z$ transformation (Brannick et al., 2019; Field, 2005; Hafdahl & Williams, 2009, Hunter & Schmidt, 2015) and that the Fisher $z$ transformation is often applied in practice.

Using Pearson's $r$ for meta-analysis when interested in heterogeneity has the advantage that resulting estimates on this scale are more interpretable that on the Fisher $z$ scale. Although Fisher $z$ correlations can be back-transformed into Pearson's $r$ for interpretation, its heterogeneity cannot. A workaround is to compute a prediction interval and convert this interval onto the product-moment scale (REF). The width of the interval can then be used as an estimate of the heterogeneity. We do not believe this increases interpretability in our case and so prefer to directly estimate the heterogeneity. However, the disadvantage to using Pearson's $r$ when estimating heterogeneity is that 1) the effect size is bounded to {-1, 1} and 2) the sampling variance $$ depends on the effect size. The bounded nature of the Pearson correlation means that large levels of heterogeneity will lead to substantial truncation of effect sizes if the average effect size is also large. That the sampling variance covaries with effect size leads to a small positive bias in heterogeneity estimates (supplement XX). This can be solved by replacing $\rho$ with the average correlation across studies $\bar{r}$ (Hunter & Schmidt, 2015), and removing this covariance is also the primary intention behind the Fisher $z$ transformation. Fisher $z$ correlations also have the advantage that they are not bounded to {-1, 1}. A disadvantage to using Fisher $z$ correlations, in addition to interpretability of heterogeneity estimates, is that they require first backtransforming into product-moment correlations before adding measurement error, and then transforming to Fisher $z$ again for meta-analysis. Especially for large Fisher $z$ values (which can happen if the average effect size and heterogeneity are large) this transformation process may introduce inaccuracies in the estimates. Given the common use of both effect size types and their mix of advantages and disadvantages we report results for both.

@@

We apply two meta-analytic models to ensure that our results are not model-dependent: the Hedges and Vevea (1998) random-effects  model and the Hunter and Schmidt (2015) 'bare-bones' random-effects model. The Hedges and Vevea (HV) model uses inverse variance weight computed as $w_i = 1 / (V_i + T^2)$ (e.g., Borenstein et al., 2010) where $V_i$ is the sampling variance of study $i$ and $T^2$ is the estimated between-studies variance. The Hunter and Schmidt (HS) model weighs studies by sample size $w_i = N_i$. The differences between these weights are relatively minor (Borenstein et al., 2010; Hunter and Schmidt, 2015, p. XX). Although the HV model is typically applied to Fisher z transformed correlations (e.g., Borenstein et al., p. 41-43), we apply it to both transformed and untransformed correlations for the reasons described in the previous section. The HS model typically includes correcting for unreliability or other measurement artifacts. However, we implement it without corrections ('bare-bones'; Hunter & Schmidt, 2015) because this allows us to focus on the bias in heterogeneity estimates rather than the performance of corrections.

Estimating heterogeneity in both the HV and bare-bones HS model essentially consists of comparing the observed variance in effect sizes with what would be expected from sampling error alone and then ascribing any excess observed variance to heterogeneity (Borenstein et al., 2010, Box 1; Hunter & Schmidt, 2015, p. XX). However, the process for doing so differs somewhat between methods. We apply the Restricted Maximum Likelihood (REML) estimator of heterogeneity when estimating heterogeneity in the Hedges and Vevea (1998) model, because it is the generally recommended heterogeneity estimator (Veroniki et al., 2016; Viechtbauer, 2005). Describing the REML estimator in detail is beyond the scope of this paper and we refer interested readers to the description in Viechtbauer (2005). The Hunter and Schmidt (2015) model estimates heterogeneity as $\tau^2 = \sigma^2_r - \sigma^2_e$ where  $\sigma^2_r$ is the observed variance across correlations and $\sigma^2_e$ is the average sampling error variance across studies, computed as $ \frac{\Sigma[N_i \sigma^2_{ei}]}{\Sigma N_i}$ (p. 87, 2nd edition) and $\sigma^2_{ei}$ is the usual sampling variance for the Pearson correlation but with $\rho$ replaced with the sample size weighted average correlation $\bar{r} = \frac{\Sigma n_i r_i}{\Sigma n_i}$ instead of $r_i$ as is common. A generalized version of the HS estimator for effect sizes other than correlations can be found in Viechtbauer (2005; 2015) (see ?rma).

>>>>

Perhaps the most common way to meta-analyze correlations is to first transform them into Fisher $z$ correlations (Borenstein et al., 2009, p. 41 - 43). The Fisher $z$ transformation is the inverse hyperbolic tangent of the product moment correlation and can be computed as $z = 0.5 \times ln\left(\frac{1+r}{1-r}\right)$ and its variance approximated as $V_z = 1 / (n -3)$. There is controversy in the use of the Fisher $z$ transformation for meta-analysis (e.g., Hunter & Schmidt, 2015, p. XX, ch. 5 "Use of Fisher’s z in Meta-Analysis of Correlations"), largely related to whether Pearson's $r$ or Fisher's $z$ leads to less bias in average correlations (Schulze, 2007). However, estimating heterogeneity does not seem to have been the focus of most of the literature's discussion of the appropriateness of the Fisher $z$ transformation (Field, 2005; Hafdahl & Williams, 2009, Hunter & Schmidt, 2015). Hall and Brannick (2002) do report heterogeneity estimates but focus on the coverage of prediction intervals ('credibility intervals' in Hunter & Schmidt terminology) which combine the heterogeneity estimate and point estimate, and in Brannick et al., (2019) they examine results only after corrections for attenuation. Given the inconclusive evidence and that the Fisher $z$ transformation is often applied in practice, we use it as one of the effect size types in our simulations.

The other effect size type we us is Pearson's $r$. Pearson's $r$ has the advantage that resulting heterogeneity estimates on this scale are more interpretable that on the Fisher $z$ scale. Although Fisher $z$ correlations can be back-transformed into Pearson's $r$ for interpretation, its heterogeneity cannot. A workaround is to compute a prediction interval and convert this interval onto the product-moment scale (Hedges & Vevea, 1998). The width of the interval can then be used as an estimate of the heterogeneity. We prefer to compute comparable heterogeneity values for Pearson's $r$ and Fisher's $z$ (see the methods section 'Parameter values') and report the bias in heterogeneity estimates, as we believe this is more interpretable for applied meta-analysts.

The disadvantage to using Pearson's $r$ when estimating heterogeneity is that 1) the effect size is bounded to {-1, 1} and 2) the sampling variance $\sigma^2_r = \frac{(1-\rho^2)^2}{n - 1}$ depends on the true effect size $\rho$. The bounded nature of the Pearson correlation means that large levels of heterogeneity will lead to substantial truncation of effect sizes if the average effect size is also large. That the sampling variance covaries with effect size leads to a small positive bias in heterogeneity estimates (supplement XX). This can be solved by replacing $\rho$ with the average correlation across studies $\bar{r}$ (Hunter & Schmidt, 2015), or using the Fisher $z$ transformation. Fisher $z$ correlations have the additional advantage that they are not bounded to {-1, 1}. A disadvantage to using Fisher $z$ correlations, in addition to interpretability of heterogeneity estimates, is that they require backtransforming into product-moment correlations before adding measurement error, and then transforming to Fisher $z$ again for meta-analysis. Especially for large Fisher $z$ values (which can happen if the average effect size and heterogeneity are large) this transformation process may introduce inaccuracies in the estimates. Given the common use of both effect size types and their mix of advantages and disadvantages we report results for both.

@@

We apply two meta-analytic models to ensure that our results are not model-dependent: the Hedges and Vevea (1998) random-effects  model and the Hunter and Schmidt (2015) 'bare-bones' random-effects model. The Hedges and Vevea (HV) model uses inverse variance weight computed as $w_i = 1 / (V_i + T^2)$ (e.g., Borenstein et al., 2010) where $V_i$ is the sampling variance of study $i$ and $T^2$ is the estimated between-studies variance. The Hunter and Schmidt (HS) model weighs studies by sample size $w_i = N_i$. The differences between these weights are relatively minor (Borenstein et al., 2010; Hunter and Schmidt, 2015, p. XX). Although the HV model is typically applied to Fisher $z$ transformed correlations (e.g., Borenstein et al., p. 41-43), we apply it to both transformed and untransformed correlations to enable a more direct comparison with the HS model and facilitate interpretation. The HS model typically includes correcting for unreliability or other measurement artifacts. However, we implement it without corrections ('bare-bones'; Hunter & Schmidt, 2015) because this allows us to focus on the bias in heterogeneity estimates rather than the performance of corrections.

Estimating heterogeneity in both the HV and bare-bones HS model essentially consists of comparing the observed variance in effect sizes with what would be expected from sampling error alone and then ascribing any excess observed variance to heterogeneity (Borenstein et al., 2010, Box 1; Hunter & Schmidt, 2015, p. XX). However, the process for doing so differs somewhat between methods. We apply the Restricted Maximum Likelihood (REML) estimator of heterogeneity when estimating heterogeneity in the Hedges and Vevea (1998) model, because it is the generally recommended heterogeneity estimator (Veroniki et al., 2016; Viechtbauer, 2005). Describing the REML estimator in detail is beyond the scope of this paper and we refer interested readers to the description in Veroniki et al., (2016). The Hunter and Schmidt (2015) model estimates heterogeneity as $\tau^2 = \sigma^2_r - \sigma^2_e$ where  $\sigma^2_r$ is the observed variance across correlations and $\sigma^2_e$ is the average sampling error variance across studies, computed as $ \frac{\Sigma[N_i \sigma^2_{ei}]}{\Sigma N_i}$ (p. 87, 2nd edition) and $\sigma^2_{ei}$ is the usual sampling variance for the Pearson correlation but with $\rho$ replaced with the sample size weighted average correlation $\bar{r} = \frac{\Sigma n_i r_i}{\Sigma n_i}$ instead of $r_i$ as is common. A generalized version of the HS estimator for effect sizes other than correlations can be found in Viechtbauer (2005; 2015).
